An Implementation of the Active Contours without Edges model and the Logic Framework for Active contours on multi-channel images

### 2010-12-13 cs870/project/Report/Paper.tex
need to fix errors in refs

\usepackage{cvpr}
\usepackage{cite}
\usepackage{url}
\usepackage{graphicx}
\usepackage{clrscode}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{array}
\usepackage{color}
\usepackage{balance}

@@

\begin{abstract}
%\boldmath
The abstract goes here.
\end{abstract}

\section{Introduction}
Image segmentation or boundary detection is a very important problem in the area of Image Processing, and has received a lot of attention in the past. The
classical Active Contour model (or Snakes) proposed by Kass et al.~\cite{kass1988snakes} was the first model to use the idea of energy minimization to attract
a contour to the edges of the objects in an image. The Snakes model was very successful and variations of it very highly adopted later on
(E.g.~\cite{caselles1997geodesic}. Most of these models used the level set formulation for propagating fronts to evolve the curve. However, these models highly
depended on curvature motion (motion defined by the gradient of the curve) which led to poor performance in smoothed edges. To overcome these limitations, Chan
and Vese~\cite{chan2001active} propose an active contour model that does not depend on the edges (i.e the gradient) for propagating the curve to detect the
boundary of the object. Instead, they use a region-based approach based on the Mumford-Shah model~\cite{mumford1989optimal} to divide the image
into two regions: one inside the propagating curve, and one outside. The curve is at the boundary of the object if there is no difference in intensities inside
the curve as well as outside the curve.

@@

\begin{figure}[t]
\centering
\includegraphics[width=7cm]{sc_unionsimple.png}
\caption{Simple union example. Size = 300 x 300, $\phi_{0}(x,y) = - \sqrt{(x - 150)^2 + (y - 150)^2)} + 75$,  no reinitialization, cpu = 1.32 s, 
iterations = 2.}
\label{fig:sc_unionsimple}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=7cm]{sc_intersectionsimple.png}
\caption{Simple intersection example. Size = 300 x 300, $\phi_{0}(x,y) = - \sqrt{(x - 150)^2 + (y - 150)^2)} + 75$,  no reinitialization, cpu = 1.83 s, 
iterations = 2.}
\label{fig:sc_intersectionsimple}
\end{figure}

@@

There are mainly three parameters in this model that can be varied: $\mu, \lambda_{1}, \lambda_{2}$. We would usually want both $\lambda_{1}$ and $\lambda_{2}$
to be equal to 1 to indicate that we care both about the difference of intensities inside and outside. A smaller weighting for one of them would mean that we
would ignore some of the variances in intensities in this area. A larger weighting means that we are magnifying the difference in intensities in that area
(i.e. we want to detect any minor changes). We show an example of these variations in Figure~\ref{fig:cv_eg12} where we vary
$lambda_1$ and $\lambda_2$. When we set $\lambda_1 = 0.2$ and $\lambda_2 = 1$, less details within the interior of the brain is detected. When
we set $\lambda_1 = 2$ and $\lambda_2 = 1$, we can see that more details within the brain are detected. When we set $\lambda_1 = 5$, and $\lambda_2 = 0.01$ (a
very extreme case), we can see that parts of outside boundaries are not well detected, while many details are detected within the brain.

@@

\begin{figure*}[t]
\centering
\includegraphics[width=12cm]{cv_eg12.png}
\caption{Effect of varying $\lambda_1$ which controls how much details are detected inside the contour. The figure shows the final segmentation in each case.}
\label{fig:cv_eg12}
\end{figure*}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=12cm]{sc_noisy.png}
\caption{Same image as Figure~\ref{fig:threech}, but with 'salt \& pepper' noise with a variation of 0.1. Decreasing $\lambda$ from 255*255 to 125 allowed less
noise to be detected. Size = 200 x 200, $\phi_{0}(x,y) = - \sqrt{(x - 100)^2 + (y - 100)^2)} +
50$, no reinitialization. Cpu = 3.17 sec and iterations = 5 when $\lambda = 255*255$. Cpu = 146.46 sec and iterations = 125 when $\lambda =
125$.}
\label{fig:sc_noisy}
\end{figure*}

@@

Unfortunately, the complement did not work properly in our implementation. We closely followed the definitions in the original paper where $z_i^{in'} = 1 -
z_i^{in}$ and $z_i^{out'} = 1 - z_i^{out}$. The results was actually successfully detected, but continuous flickering occurred, and the curve did not stop in
one position. We discuss this results in more details in Section~\ref{sec:disc}.

@@

\section{Difficulties and Discussion}
\label{sec:difficulties}

\section{Conclusion}
\label{sec:concl}
The conclusion goes here.

\bibliographystyle{IEEEtran}

\bibliography{references}

>>>>

\usepackage{cvpr}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage[pagebackref=true,breaklinks=true,letterpaper=true,colorlinks,bookmarks=true]{hyperref}

@@

\begin{abstract}
In this project, we provide an implementation for the Chan-Vese model for active contours without edges, and the Sandberg-Chan logic framework for active
contours on multi-channel images. The Chan-Vese model is a special case of the more general Mumford-Shah functional for segmentation and level sets. It differs
from other active contour modesl in that it does not depend on the gradient of the contour ($\phi$), therefore it is more capable of detecting objects whose
boundaries may not be defined by a gradient (e.g. blurry images, noisy images). The Sandberg-Chan model builds on top of the Chan-Vese model by allowing the
contour to evolve simultaneously on multiple channels. It also offers a set of logic operations that can be applied to those channels (union, intersection,
complement, and their combinations). We evaluate our implementation based on several criteria, such as: independence of initial curve position, successful
detection of holes, performance with blurred and noisy images, and response to scaling parameters.
\end{abstract}

\section{Introduction}
Image segmentation or boundary detection is a very important problem in the area of Image Processing, and has received a lot of attention in the past. The
classical Active Contour model (or Snakes) proposed by Kass et al. \cite{kass1988snakes} was the first model to use the idea of energy minimization to attract
a contour to the edges of the objects in an image. The Snakes model was very successful and variations of it very highly adopted later on
(E.g.~\cite{caselles1997geodesic}. Most of these models used the level set formulation for propagating fronts to evolve the curve. However, these models highly
depended on curvature motion (motion defined by the gradient of the curve) which led to poor performance in smoothed edges. To overcome these limitations, Chan
and Vese~\cite{chan2001active} propose an active contour model that does not depend on the edges (i.e the gradient) for propagating the curve to detect the
boundary of the object. Instead, they use a region-based approach based on the Mumford-Shah model \cite{mumford1989optimal} to divide the image
into two regions: one inside the propagating curve, and one outside. The curve is at the boundary of the object if there is no difference in intensities inside
the curve as well as outside the curve.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6cm]{sc_unionsimple.png}
\caption{Simple union example. Size = 300 x 300, $\phi_{0}(x,y) = - \sqrt{(x - 150)^2 + (y - 150)^2)} + 75$,  no reinitialization, cpu = 1.32 s, 
iterations = 2.}
\label{fig:sc_unionsimple}
\end{figure}

\begin{figure}[t!]
\centering
\includegraphics[width=6cm]{sc_intersectionsimple.png}
\caption{Simple intersection example. Size = 300 x 300, $\phi_{0}(x,y) = - \sqrt{(x - 150)^2 + (y - 150)^2)} + 75$,  no reinitialization, cpu = 1.83 s, 
iterations = 2.}
\label{fig:sc_intersectionsimple}
\end{figure}

@@

There are mainly three parameters in this model that can be varied: $\mu, \lambda_{1}, \lambda_{2}$. We would usually want both $\lambda_{1}$ and $\lambda_{2}$
to be equal to 1 to indicate that we care both about the difference of intensities inside and outside. A smaller weighting for one of them would mean that we
would ignore some of the variances in intensities in this area. A larger weighting means that we are magnifying the difference in intensities in that area
(i.e. we want to detect any minor changes). We show an example of these variations in Figure~\ref{fig:cv_eg12} where we vary
$\lambda_1$ and $\lambda_2$. When we set $\lambda_1 = 0.2$ and $\lambda_2 = 1$, less details within the interior of the brain is detected. When
we set $\lambda_1 = 2$ and $\lambda_2 = 1$, we can see that more details within the brain are detected. When we set $\lambda_1 = 5$, and $\lambda_2 = 0.01$ (a
very extreme case), we can see that parts of outside boundaries are not well detected, while many details are detected within the brain.

@@

\begin{figure}[t]
\centering
\includegraphics[width=7cm]{cv_eg12.png}
\caption{Effect of varying $\lambda_1$ which controls how much details are detected inside the contour. The figure shows the final segmentation in each case.}
\label{fig:cv_eg12}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[width=7cm]{sc_noisy.png}
\caption{Same image as Figure~\ref{fig:threech}, but with 'salt \& pepper' noise with a variation of 0.1. Decreasing $\lambda$ from 255*255 to 125 allowed less
noise to be detected. Size = 200 x 200, $\phi_{0}(x,y) = - \sqrt{(x - 100)^2 + (y - 100)^2)} +
50$, no reinitialization. Cpu = 3.17 sec and iterations = 5 when $\lambda = 255*255$. Cpu = 146.46 sec and iterations = 125 when $\lambda =
125$.}
\label{fig:sc_noisy}
\end{figure}

@@

Unfortunately, the complement did not work properly in our implementation. We closely followed the definitions in the original paper where $z_i^{in'} = 1 -
z_i^{in}$ and $z_i^{out'} = 1 - z_i^{out}$. The results was actually successfully detected, but continuous flickering occurred, and the curve did not stop in
one position. We discuss this results in more details in Section~\ref{sec:difficulties}.

@@

\section{Difficulties and Discussion}
\label{sec:difficulties}
We have faced many challenges throughout the implementation of this project. Some of the challenges caused major delays in the implementation cycle and hindered
us from improving our code and adding more features to it.

\subsection{Varying $\mu$}
$\mu$ is a scaling factor that is used in both the Chan-Vese model and the Sandberg-Chan model to control the weight given to the length of the contour.
Therefore, varying the value of $\mu$ will affect the ability to segment objects in noisy images, especially when the noise is similar to Matlab's \textit{salt
and pepper} noise. However, varying $\mu$ in our implementation does not really have any impact on the length of the contour. Figure [\ref{fig:cvmu}] shows the
segmentation of an object in a noisy for different values for $\mu$. The resulting segmentation is the same for all values of $\mu$. It is worth noting that
this challenge kept us busy for over two days, after which we decided we should move on to other parts of the project.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{cvmu.png}
\caption{$\mu$ does not have any effect on the length of the contour. \textit{Salt and pepper} noise was added to the original image for testing
purposes. Top: original image with noise. Bottom: segmentation for various values of $\mu$.}
\label{fig:cvmu}
\end{figure*}

\subsection{Computing the complement}
The Sandberg-Chan model supports the complement logic operator, so one can segment the logic object ($A \cap \neg B$) (i.e. segment what is in channel A and not
in channel B). The model defines the complement of the object in channel \textit{i} as follows:

\begin{equation}
\label{eqn:complement}
\begin{align*}
{z_{i}}^{in^{'}} = 1 - {z_{i}}^{in}, \\
{z_{i}}^{out^{'}} = 1 - {z_{i}}^{out}.
\end{align*}
\end{equation}

Although the definition is very simple, it does not work in practice. We implemented the complement logic operator as defined in Equation
[\ref{eqn:complement}], and the segmentation does not stop at the required result. It rather keeps evolving after reaching the required solution, then keeps
flipping between two segmentations: the correct segmentation, and the segmentation of $A \cap B$. Figure [\ref{fig:sccomp}] shows this scenrio. We tried to
debug the reasons for this problem for three days.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{sccomp.png}
\caption{The complement operator causes the contour to go through an infinite loop between two segmentations: the correct segmentation, and the segmentation
of $A_1 \cap A_2$. Bottom: correct and incorrect segmentation respectively}
\label{fig:sccomp}
\end{figure*}

\subsection{Defining the initial contour ($\phi_0$)}
Both the Chan-Vese model and the Sandberg-Chan model defines the initial contour $\phi_0$ as a signed distance function where $\phi_0$ is positive inside the
contour $C$, zero on the boundary of $C$ ($\partial C$), and negative outside the contour $C$. Figure [\ref{fig:phinote}] illustrates those assumptions.

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{phinote.png}
\caption{Assumptions about the way $\phi_0$ is defined.}
\label{fig:phinote}
\end{figure}

Since we model our initial contour as a circle, we have to modify the regular equation of the circle to match those assumption. Therefore, we initialize
$\phi_0$ as follows:

\begin{equation}
\label{eqn:phinote}
\phi_{0} (x, y) = radius - \sqrt{ (x - c_x)^2 + (y - c_y)^2 },
\end{equation}
where $c_x$ and $c_y$ are the x-coordinate and y-coordinate of the center of the circle respectively.

Although this definition of $\phi_0$ is compatible with most of the images we used to test our implementation, it did fail to correctly segment the objects in
some images. In fact, the results were the opposite of the used logic operator (i.e. the output is the intersection of the channels, while it should be the
union). After spending quite sometime to resolve this problem and checking what might be the source of the error, we tried using the regular equation of a
circle to initialize $\phi_0$ (Equation [\ref{eq:phinotenormal]}). We were very intrigued to notice that this modification resulted in the correct segmentation
of the input channels. Figure [\ref{fig:phidef}] shows the final segmentation using both definitions for $\phi_0$.

\begin{equation}
\label{eqn:phinotenormal}
\phi_{0} (x, y) = \sqrt{ (x - c_x)^2 + (y - c_y)^2 } - radius,
\end{equation}
where $c_x$ and $c_y$ are the x-coordinate and y-coordinate of the center of the circle respectively.

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{phidef.png}
\caption{}
\label{fig:phidef}
\end{figure*}

\subsection{Mistyped Equations}
As previously mentiond in Section [\ref{sec:sandberg-chan}], there is a typo in the equations that define $z_in$ and $z_out$ in the Sandberg-Chan model
(equation 3 in thier paper). In addition, they have another typo in the expansion of the the Eular-Lagrange equation for the PDE that models the problem. Both
typos consumed over a week of our time until we were able to find out what are the correct equations in both cases. We could have spent this time overcoming
other challenges that we faced while implementing both models.

\section{Conclusion}
\label{sec:concl}
In this project, we have provided an implementation for the Chan-Vese model for active contours without edges and the Sandberg-Chan logic framework for
multi-channel images, both written in Matlab. Our experiments have shown that the Chan-Vese model is effective for various types of images. In specific, it is
useful when an edge-based segmentation is not applicable (i.e. edges do not depend on a gradient). However, our implementation fails to respond to different
values for the contour length scaling parameter, $\mu$. This affects the ability of our implementation to handle Matlab's \textit{salt \& pepper} noise. On the
other hand, changing the values for the intensity weight parameters $\lambda_1$ and $\lambda_2$ affects the amount of details obtained inside and outside the
segmented object respectively.

We have also shown that our implementation for the Sandberg-Chan logic framework handles different logic combinations for multiple channels. We have also
experimented using it as a segmentation tool for colored images by modeling an input colored images as a 3-channel image (red, green, blue). However, our
implementation for the \textit{complement} logic operation seems to be missing a hidden trick in the Sandberg-Chan logic framework.

\bibliographystyle{IEEEtran}
\bibliography{references}
